---
title: "Day 1"
author: "Caio Kenup"
date: "2023-07-11"
output: html_document
---

# Introduction

Data visualization plays a crucial role in post-graduate research and analysis. In this primer, we will explore the powerful plotting capabilities of `ggplot` in R. `ggplot` is a widely used package that allows you to create stunning and customizable visualizations. By the end of this primer, you will have a solid foundation in creating compelling plots using `ggplot`.

## Basic Plotting Concepts with ggplot

To get started, we need to install and load the `ggplot2` package. If you haven't installed it yet, you can do so by running the following code:

```{r}
require(ggplot2)
```


# Simple Scatterplot

For this example we will use the sample dataset `cars`


```{r}
str(cars)
```



This is how you would do a scatterplot in base R

```{r}
plot(x=cars$speed,y=cars$dist,xlab = "Speed",ylab="Dist")
```

And this is how you have the same scatterplot in ggplot2

```{r}
ggplot(data=cars,mapping = aes(x = speed,y=dist))+
  geom_point()
```


`geom_point()` is a function o the _geometry_ family - these are the functions that "draw" stuff on your plot. Some of them are

- `geom_point()`
- `geom_line()`
- `geom_bar()`
- `geom_ribbon()`

Now let's take a more complex dataset, `mtcars`

```{r}
str(mtcars)
```


#


```

```